Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-06-26 Text/Response_to_referee1.txt

This document contains the authors responses to the initial referee's report for ApJL submission LET26384. Several
changes have been made to the manuscript. The most significant changes are an expanded discussion section and an
additional figure, figure 4, depicting the aspect ratio of the disks at different times. Detailed responses to the
referee's comments follow below.

1. "Even though there are some interesting results in the paper, most of the major results on migration in
gravitationally unstable disks were already reported in Baruteau et al. (2011). There is therefore no need for fast
publication as a Letter, especially since there appears to be very little explanation of the results in the current
manuscript. I would encourage the authors to resubmit a more detailed version of this work to the main journal, in
particular focusing on the differences of their approach compared to that of Baruteau et al. (2011)."

We believe it is important to note that the paper referenced here by the referee had been accepted to the MNRAS only one
week before our submission. It appears that the article is currently "in press" and is only publicly available on
astro-ph. As such we feel it should not block the publication of this article to the ApJL. In addition, as outlined in
our extended comments in the discussion section, Baruteau et al. use very different physics in their protoplanetary disk
simulations. The simulations presented in our work are fully three dimensional and use realistic radiative physics,
while the Baruteau et al. simulations are two dimensional and use an idealized cooling prescription. It has been shown
by many authors that both 3D and realistic cooling are required to accurately characterize and measure GIs.

2. "The critical dimensionless parameter governing the linearity of planet-disk interactions is q/h^3, where q is the
mass ratio of the planet - central star, and h is the aspect ratio of the disk. A plot of h(r) would therefore be
helpful."

Disk aspect ratios for the end of both planet runs and the corresponding time in the fiducial run have been provided in
figure 4.

3. "Gap formation is governed by above parameter together with viscosity (Crida et al. 2006). In the current set-up,
gravitational and Reynolds stresses probably strongly depend on radius. A radial plot of the stresses would give some
insight in gap formation at various locations in the disk."

Although we would like to include such figures and relevant discussion, and will in our future more detailed article,
the constraints of the Letter format prevent us from doing so.

4. "If the planet does not open up a gap, it is subject to Type I migration. The migration rate can then be very
sensitive to background temperature and density profiles (Paardekooper et al. 2010). Plots of azimuthally averaged
surface density and temperature profiles would help to determine whether one would expect inward or outward Type I
migration."

In fact, one of the main points of the article is that distinctions such as type II, type II, and type III migration,
which depend on the presence and character of a gap in a laminar disk, need not apply in a GI active disk. Although the
overall migration rate is similar to that predicted by type I migration theory, the same mechanism is not acting here.

5. "The authors state that they do not resolve the width w of the horseshoe region of the planet. Again, the critical
parameter is q/h^3. For h=0.1, w should already be larger than the scale height. Even if h=0.2, w should be close to
h/2. Not resolving w therefore implies not resolving h, which would be quite bad. Are the authors sure they can not
resolve w?"

We have removed the statement which led the referee to this conclusion. What we intended to convey was the fact that we
lack sufficient resolution to accurately measure corotation torques, not that we could not resolve the horseshoe region.

6. "In the introduction, please note that Baruteau & Masset (2008) were actually the first to include self-gravity in
migration simulations (but in laminar disks only). It is the gravitationally unstable nature of the disk that is new in
the present study and in Baruteau et al. (2011)."

As this article deals exclusively with GI active disks, and we did not indicate that another article was the first to
include self-gravity we declined to cite Baruteau & Masset (2008).

7. "I find the 't=0 planet run' difficult to understand. The authors seem to claim that the presence of the planet
affects the development of GI. How can they justify these initial conditions? If the planet was formed by GI, the disk
should already be turbulent. If it formed in a laminar disk (or in a previous GI burst), it should already have migrated
inward."

We have include the following text to illuminate some of the possible scenarios: "This run investigates the effects of
an existing planet on the growth of GIs, which could occur if disk the disk undergoes episodic or localized
instabilities."

8. "I am not sure what the role of the ILR of the dominant spiral mode is in stopping the migration of the
planet. Intuitively, one might expect an ILR to push the planet inward. What does the torque balance look like?"

Again, space limitations prevent us from presenting full torque diagnostics, however they will be presented in our
follow-up article. As we say in the article, this is an area for further research.

9. "The interaction of the planet with all the GI modes appears to be quite stochastic. It seems to me that very little
can be concluded from a single run."

The simulations presented here are by no means comprehensive, but are intended to give an indication of what possible
effects GIs might have on migration, and vice-versa. We provide as much detail for the given runs as allowed in the
Letter format, including additional planet runs (as will be done in the follow-up article) is not feasible in the
allowed space.

### 2011-07-12 Text/migration.tex
Responded to second referee report, should be the final submission.
	new file:   Response_to_referee2.txt
	modified:   general.bib
	modified:   migration.tex
	modified:   migration_apjlsubmission.tex

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the
architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational
interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state
(EOS), and the self-gravity of the gas disk have usually been ignored; the effects of a non-isothermal EOS have only
recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. Emerging studies, such as
\citet{baruteau2011} and the one reported here, are beginning to address some of these issues.  \citet{boss2005} and
\citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks
were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim
0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur
\citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into
clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively
massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU)
\citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from
semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part
on their radial migration, which is a chaotic and messy affair in a fragmenting disk
\citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities (GIs) may be
episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of
young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves,
such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

To improve our understanding of how planets migrate in GI-active disks that are not fragmenting, we have begun a
systematic study, using numerical 3D radiative hydrodynamics, where we investigate the effects on both the disk and the
planet of inserting a planet-mass object into disks susceptible to GIs. Using techniques developed in earlier research
by our group \citep{pickett2003, mejia2005,cai2006,cai2008,boley2006,boley2007b,michael2010}, we can identify the
dominant spiral waves in a simulation and analyze how the waves interact with the planet's motion. Our goal is to
determine both the effect of giant planets on GIs and the effect of GIs on planet migration. Because GIs are sensitive
to radiative physics, we use a well-tested radiative scheme \citep{boley2007b} and realistic opacities
\citep{dalessio2001}.

@@

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too
different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3
ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae
that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from
the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation (see figure \ref{fig:hr} for
disk aspect ratios). While the interaction with a GI-active disk does not seem to result in a drastically different
overall migration rate, the direction and magnitude of migration in both simulations is quite variable and seems to
depend strongly on interactions of the planet with discrete modes in the disk. In $t = 10$ planet run, the overall
radial migration rate is similar but is even more erratic in direction than in the $t = 0$ planet run. The important
departures from laminar disk theory are threefold: the more erratic nature of migration, the importance of interactions
with discrete low-order modes, and stalling of migration near the inner edge of the strongest GI activity. The migration
is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons,
it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its
placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes
(for $t = 10$ type runs).

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research
was conducted with the support of NASA grants from the Origins of Solar Systems Program (NNG05GN11G and
NNX08AK36G). A.C.B. is currently a Sagan Postdoctoral Fellow at U. Florida. This material is based upon work supported
by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258, OCI-0504075, and
CNS-0521433, and through TeraGrid resources provided by Indiana University, PSC, SDSC, and NCSA.

>>>>

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the
architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational
interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state
(EOS), and the self-gravity of the gas disk have usually been ignored; the effects of a non-isothermal EOS have only
recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. Emerging studies, such as
\citet{baruteau2011} and the one reported here, are beginning to address some of these issues.  \citet{boss2005} and
\citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks
were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim
0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur
\citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into
clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively
massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU)
\citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from
semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part
on their radial migration, which is a chaotic and messy affair in a fragmenting disk
\citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities may be
episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of
young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves,
such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

To improve our understanding of how planets migrate in GI-active disks that are not fragmenting, we have begun a
systematic study to investigate the effects on both the disk and the planet by inserting a planet-mass object into disks
susceptible to GIs. Using techniques developed in earlier research by our group \citep{pickett2003,
  mejia2005,cai2008,boley2006,boley2007b,michael2010}, we can identify the dominant spiral waves in a simulation and
analyze how the waves interact with the planet's motion. Our goal is to determine both the effect of giant planets on
GIs and the effect of GIs on planet migration. Because GIs are sensitive to radiative physics, we use a well-tested
radiative scheme \citep{boley2007b} and realistic opacities \citep{dalessio2001}.

@@

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too
different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3
ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae
that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from
the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation (see figure \ref{fig:hr} for
disk aspect ratios). While the interaction with a GI-active disk does not seem to result in a drastically different
overall migration rate, the direction and magnitude of migration in both simulations is quite variable and seems to
depend strongly on interactions of the planet with discrete modes in the disk. In $t = 10$ planet run, the overall
radial migration rate is similar but is even more erratic in direction than in the $t = 0$ planet run. The important
departures from laminar disk theory are threefold: the more erratic nature of migration, the importance of interactions
with discrete low-order modes, and stalling of migration near the inner edge of the strongest GI activity. The migration
is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions, similar to the erratic
migration seen in planet/disk simulations with Rossby vortex instabilities \citep{yu2010}. For all these reasons,
it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its
placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes
(for $t = 10$ type runs).

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research
was conducted with the support of NASA grants from the Origins of Solar Systems Program (NNG05GN11G and
NNX08AK36G). A.C.B. is currently a Sagan Postdoctoral Fellow at U. Florida. This material is based upon work supported
by the National Science Foundation under Grants No. OCI-0451237, OCI-0535258, OCI-0504075, and through TeraGrid resources provided by Indiana University, PSC, and NCSA.

### 2011-07-12 Text/migration_apjlsubmission.tex
Responded to second referee report, should be the final submission.
	new file:   Response_to_referee2.txt
	modified:   general.bib
	modified:   migration.tex
	modified:   migration_apjlsubmission.tex

\bibitem[{{Cai} {et~al.}(2006){Cai}, {Durisen}, {Michael}, {Boley},
  {Mej{\'{\i}}a}, {Pickett}, \& {D'Alessio}}]{cai2006}
{Cai}, K., {Durisen}, R.~H., {Michael}, S., {Boley}, A.~C., {Mej{\'{\i}}a},
  A.~C., {Pickett}, M.~K., \& {D'Alessio}, P. 2006, ApJL, 636, L149

@@

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/amt.eps}
\caption{(Color figure available in the online version) Global amplitudes of nonaxisymmetric density perturbations $A_m$
  as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of
  \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge
  effects in the inner disk. The top and bottom panels correspond to the fiducial run (no planet) and the $t = 0$ planet
  run, respectively.}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planeta.eps}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The
  dashed-diamond and solid-star lines (colored red and black in the online version) indicate the curves ($\varpi(t)$),
  points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The
  symbols are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as
  explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planete.eps}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in
  the text, as a function of time. The diamonds and stars (colored red and black in the online version) indicate the
  points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/hr_migration.eps}
\caption{(Color figure available in the online version) Plots of the disk aspect ratio $h/r$ as determined by a decrease
  from midplane density by a factor of $e^2$ for the fiducial and $t=0$ runs (black) at $10.5$ ORP and the fiducial and
  $t=10$ runs (light gray) at $20.5$ ORP.}
\label{fig:hr}
\end{figure}

>>>>

@@

\bibitem[{{Yu} {et~al.}(2010){Yu}, {Li}, {Li}, {Lubow}, \& {Lin}}]{yu2010}
{Yu}, C., {Li}, H., {Li}, S., {Lubow}, S.~H., \& {Lin}, D.~N.~C. 2010, ApJ,
  712, 198

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{amt.eps}
\caption{(Color figure available in the online version) Global amplitudes of nonaxisymmetric density perturbations $A_m$
  as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of
  \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge
  effects in the inner disk. The top and bottom panels correspond to the fiducial run (no planet) and the $t = 0$ planet
  run, respectively.}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{planeta.eps}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The
  dashed-diamond and solid-star lines (colored red and black in the online version) indicate the curves ($\varpi(t)$),
  points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The
  symbols are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as
  explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{planete.eps}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in
  the text, as a function of time. The diamonds and stars (colored red and black in the online version) indicate the
  points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{hr_migration.eps}
\caption{(Color figure available in the online version) Plots of the disk aspect ratio $h/r$ as determined by a decrease
  from midplane density by a factor of $e^2$ for the fiducial and $t=0$ runs (black) at $10.5$ ORP and the fiducial and
  $t=10$ runs (light gray) at $20.5$ ORP.}
\label{fig:hr}
\end{figure}
